### 2012-03-03 paper_draft.tex
Reverted some stats to be congruent with pics
rewored draft with fabian

%% ** EDIT HERE **
%% PLEASE INCLUDE ALL MACROS BELOW
\long\def\authornote#1{%
        \leavevmode\unskip\raisebox{-3.5pt}{\rlap{$\scriptstyle\diamond$}}%
        \marginpar{\raggedright\hbadness=10000
        \def\baselinestretch{0.8}\tiny
        \it #1\par}}
\newcommand{\bastian}[1]{\authornote{BG: #1}}
\newcommand{\fabian}[1]{\authornote{FZ: #1}}

%% END MACROS SECTION

\begin{document}

@@

The comparison of DNA sequences from individuals in a population reveals sites which are variable. 
If differences between individuals span only over a single nucleotide then these variations are also called Single Nucleotide Polymorphisms (SNP) and different nucleotides 
at a specific site are defined as an allele. For diploid organisms, as us humans, the genotype at a given site consists of two alleles, which are independently inherited from each parent.
Different methods have been developed to read SNPs in an individual, a
process called genotyping. Microarrays are a cheap, fast and widespread used way to genotype individuals, in comparison
to whole genome sequencing. If the genotypes for more then one individual are known, the analysis of
allele frequencies can be used to associated SNPs with diseases and other inheritable traits. Genome-Wide Assocation Studies
(GWAS) make use of statistics to compare the alleles of patients to the alleles of healthy controls. This
enables GWAS to find alleles which are significantly overrepresented in patients and associate them with a disease.
This method does not allow inference of causal differences but merely identifies correlations. 
The first GWAS was published in 2005 and compared age-related macular degeneration in contrast 
to a healthy control group \cite{Klein2005}. Since the beginning, the number of participants in 
such studies has been rising: Over 1200 GWAS have been performed \cite{Johnson2009} and over 
5000 SNPs have been linked to different diseases and traits in those studies \cite{Hindorff2009}. 
%(http://www.genome.gov/page.cfm?pageid=26525384&clearquery=1#result_table)

@@

Although companies like 23andMe are willing to contribute to science, it is not easy for individual scientists to access the data. This is mainly due to concerns about privacy, liability and consent. Nevertheless, there are individual customers who are willingly sharing their data. Most do so by uploading their data to their personal website or to open software repositories like \textit{GitHub}.

This data is scattered and unorganized, making it hard to use this information. While projects like SNPedia try to keep track of all the genotyping files \cite{Cariaso2011}, they do not provide the information necessary to perform GWAS, as the phenotypic information is most often not attached to the genetic information. Projects that attach the phenotype to the genetic information, like the \textit{Personal Genome Project}, still don't allow for an easy re-use of the data, as they lack an advanced programming interface (API) or other methods by which researchers could download the data and not every customer of DTC genetic testing can participate in it.

We designed an online survey to evaluate the support in the community for a crowdsourced online plattform. Here we present the results of the survey in combination with the openSNP project. The openSNP project is an online platform, which makes it possible for DTC customers to share genotypic and phenoytypic information and connect it with external data sources. The data is made available to research via a Creative-Commones-Zero licence.

@@

\subsection*{Survey on Sharing Genetic Information}
229 people, 180 with a self-reported chromosomal sex of XY and 56 with a self-reported chromosomal sex of XX participated in the survey. The mean age of the participants is 33 (SD = 11,29). 81.7 \% reported their ethnicity as caucasian. 39.7 \% of the participants are already customers of at least one DTC genetic testing company and further 30.1 \% of them plan on becoming one in the future. 29.7 \% don't plan on becoming a DTC customer. There is no significant difference in the usage of DTC companies between chromosomal sexes (Somers-d).

@@

The openSNP project (http://opensnp.org) is an open, crowdsourced online platform for both DTC customers and researchers. DTC customers are encouraged to share their genotyping results and connect them with phenotypes to enable GWAS for researchers but the data can also be used for other research and in any other context. 
\bastian{Changed this to make it more true ;)}
\subsubsection*{Sharing genotypic information}
Currently users can upload their genotypings from the companies 23andme, deCODEme and Family Tree to/via a webinterface to the openSNP 
project. The uploaded data is published under the Creative Commons Zero-license, 
which - in accordance with the Panton Principles (doi:10.1371/journal.pbio.1001195) - 
allows a complete reuse of the data without any constraints.
Between the start of openSNP on 09/27/2011 and 27/02/2012, 286 people have signed 
up with openSNP, and 117 genotyping files were made available. The openSNP 
database lists 69,486,471 SNPs, which are distributed over 1,938,603 unique SNP-IDs.

\subsubsection*{Crowdsourcing phenotypes}
Beside uploading genotypic data, users may list new phenotypes that are not yet 
queried by openSNP. The specification of these phenotypes  is open and not limited 
to pre-defined categories. To reduce the amount of manual data curation, openSNP tries to harmonize 
the expression and spelling of the same phenotype or variation. We implemented an 
autocompletion-feature which helps users to reuse already entered phenotypes.
Users are encouraged to list as many phenotypes as possible through a simple 
achievement system which rewards users that upload their data and enter phenotypic 
information with small badges that are shown on their profile pages.

In the same timeframe that was mentioned above, all users combined have 
entered a total of 675 variations on 47 different phenotypes with those variations being the different values on a given trait or phenotype. 
\bastian{Clarification of variation/phenotype?}
See figure n for a 
distribution of data acquisition over time.
The mean number of users that have entered their variations for a single phenotype 
is 14.36 (SD 12.65), the median is 10. The distribution of how many users have 
entered their data per phenotype can be seen in figure n+1. The phenotype provided 
y the most users is the eye color, which has been entered by 54 users. There are 
two phenotypes which have so far only been provided by a single user: 
The SAT Writing score and triglyceride-levels.

\subsubsection*{Connection to external services}
In order to provide users with relevant information on their respective genotypes, openSNP scans databases of the scientific literature for specific SNPs. A total number of 15,229 documents relevant to the SNP-IDs listed in openSNP could be found in the databases of Mendeley, the Public Library of Science and SNPedia. Of the primary literature, 25 \% are released in open access journals and can be accessed free of charge by every user (Figure n+2). For usability reasons, SNPs are ranked by the amount of information gathered by the external services.The external services themselves are ranked by how easily users can access information from these sources. The SNPedia entries are given the highest impact, as those are already manually curated, followed by open access publications out of the Public Library of Science. Lowest values are given to the Mendeley results, as those aren't necessary freely available to every user. An entry on SNPedia is valued 2.5 times as high as a PLoS publication and 5 times as high as a Mendeley entry.

\subsubsection*{Data access}
OpenSNP offers extensive access to the data uploaded by users. Anyone can download single genotyping files for specific users, get archives of multiple genotyping files grouped by phenotypic variation, or access a single download that includes all genotyping files and all phenotypes in a comma-separated table. The genetic data is also accessible through the Distributed Annotation System, which offers all data for specific chromosomes and specific positions on single chromosomes. The data is additionally available over a JSON-API, which allows users to directly access data in the JSON-format. Users can discuss SNPs and phenotypes on the platform using a simple commenting system, or send each other private messages.

@@

\subsubsection*{Survey issues}
Because of this approach, the results do not reflect the general population, but overrepresent those people most likely to be interested in a project such as openSNP: customers of DTC genetic testing companies and people with a high interest in biology. As we wanted to find out about interest in crowdsourcing GWAS and about how people who had purchased DTC genetic testing felt about sharing their data with third parties, this approach was warranted.
\philipp{"Because of this approach", what approach?}

\subsubsection*{GWAS and open data}
Although prices of exome or even full genome sequencing are dropping rapidly, GWAS tend to stay considerably cheaper and offer a possibility of getting insight into genetic variations on a population level and allow an analysis of SNPs which are linked to different traits. Despite this benefit in terms of cost, it must be pointed out that GWAS can only detect correlations of SNPs with those traits and don't allow inference on the cause for any correlation. Furthermore, GWAS need a large enough sample which has to be statistically analysed by using sound methods. Nevertheless, GWAS are still frequently used and new associations still can be found (we should list some recent results here -> oxytocin maybe? we could look at the twitter-timeline of @opensnporg, i publish papers i find there).

@@

\subsubsection*{Why should a user publish their genetic information? Pros vs. Cons}
\philipp{Should this be a subsubsection, not rather a subsection?}
With openSNP, we have built a platform that can be used by customers of DTC genetic testing to easily share their genetic and phenotypic data with a wide audience, as well as by scientists and interested citizens who are looking for datasets to be used in GWAS. Customers of DTC genetic testing also benefit from an easy access to primary literature on SNPs and genetic variations they carry. While we don't have collected enough data to perform full scale GWAS yet, this might be possible in the future, as user numbers are rising. 
\subsubsection*{Privacy and health implications and ethical considerations}

As we have found in the survey on sharing such results, many of these customers are also willing to share their results with scientists and the public to help scientific progress, although people who have taken DTC genetic testing are aware of the privacy implications that come with openly sharing those results.

@@

We openly address the problem of privacy implications that come with releasing genetic data twice, during registration for openSNP and during the upload of the DTC genetic testing results. Users have to confirm that they have read and understood the disclaimer about possible side-effects that can arise by publishing their data. To further improve this process and get informed consent, we are working on implementing the procedures which are currently developed by \textit{Consent for Research} (http://www.weconsent.us).

@@

The survey included demographics such as age, chromosomal sex and ethnicity of the participants. Furthermore, it included questions on their (planned) customership of a DTC company. If they already were customers, they were also asked if they were already sharing their genetic and phenotypic data. All participants were asked if they would share their genetical or phenotypic information with their DTC company, possible answers were "Yes", "Yes, but only if they did not share my medical information with anybody else" and No".

>>>>

%% ** EDIT HERE **
%% PLEASE INCLUDE ALL MACROS BELOW
\long\def\authornote#1{%
        \leavevmode\unskip\raisebox{-3.5pt}{\rlap{$\scriptstyle\diamond$}}%
        \marginpar{\raggedright\hbadness=10000
        \def\baselinestretch{0.8}\tiny
        \it #1\par}}
\newcommand{\bastian}[1]{\authornote{BG: #1}}
\newcommand{\fabian}[1]{\authornote{FZ: #1}}
\newcommand{\philipp}[1]{\authornote{PB: #1}}
%% END MACROS SECTION
\begin{document}

@@

The availabilty of new DNA sequencing techniques has shifted the focus of biological data acuisiton towards new biomedical applications.
Many diseases – for example Alzheimers, Parkinsons or different types of cancers – are at least partly heritable and thus the genome 
of patients can be used for diagnostic purposes. Using the genetic information of patients is made possible through the sharp decrease in costs for analysing genetic information. 
Thus the importance genetic information in biomedicine will/may increase over the next years and may become a standard tool for diagnostics.

The comparison of DNA sequences from individuals in a population can reveal variable sites, which are of major interest in investigating diseases. 
A variation of only one nucleotide length at a given site is called a single Nucleotide Polymorphisms (SNP). 
The different nucleotides at this site are defined as an allele.
For diploid organisms, as us humans, the genotype at a given site consists of two alleles, which are independently inherited from each parent.
Different methods have been developed to read SNPs in an individual, a
process called genotyping. A widespread used method for genotyping individuals is the use of microarrays.
In comparison to whole genome sequencing only sites which are known to be variable are analysed, making microarrays cheaper and faster \cite{Brown1999}.

If the genotypes for more then one individual are known, the analysis of
allele frequencies can be used to associate SNPs with diseases and other inheritable traits. Genome-Wide Assocation Studies
(GWAS) make use of statistics to compare the alleles of patients to the alleles of healthy controls. This
enables GWAS to find alleles which are significantly overrepresented in patients and associate them with a disease.
This method does not allow inference of causal differences but merely identifies correlations. 
The first GWAS was published in 2005 and compared age-related macular degeneration in contrast 
to a healthy control group \cite{Klein2005}. Since the beginning, the number of participants in 
such studies has been rising: Over 1200 GWAS have been performed \cite{Johnson2009} and over 
5000 SNPs have been linked to different diseases and traits in those studies \cite{Hindorff2009}. 
%(http://www.genome.gov/page.cfm?pageid=26525384&clearquery=1#result_table)

@@

Although companies like 23andMe are willing to use the data for scientific research,
it is not easy for individual scientists to access the data. 
The generation of biomedical data by private companies raises concerns about privacy \footnote{https://www.23andme.com/about/privacy/}, 
liabilty and consent \footnote{addcitations}. 
Nevertheless, individual customers are willingly sharing their data. Most do so by uploading their data to
their personal website or to open software repositories like \textit{GitHub}. 
This data is scattered and unorganized, making it hard to use this information. While projects like SNPedia try to keep track of all 
the genotyping files \cite{Cariaso2011}, they do not provide the information necessary to perform GWAS, as the phenotypic information is 
most often not attached to the genetic information. Projects that attach the phenotype to the genetic information, 
like the \textit{Personal Genome Project}, still don't allow for an easy re-use of the data, as they lack an advanced programming interface (API) 
or other methods by which researchers could download the data and not every customer of DTC genetic testing can participate in it.

We designed an online survey to evaluate the support in the personal genetics community for a crowdsourced online plattform. Here we present the results of the survey in combination with the openSNP project. 
The openSNP project is an online platform, which makes it possible for DTC customers to share genotypic and phenoytypic information and connect it with external data sources. 
The data is made available to research via a Creative-Commones-Zero licence.

@@

\subsection*{Survey on Sharing Genetic Information}
In total 229 people, 180 with a self-reported chromosomal sex of XY and 56 with a self-reported chromosomal sex of XX participated in our survey on sharing genetic information with the public. 
The mean age of the participants is 33 (SD = 11,29). 81.7 \% reported their ethnicity as caucasian. 39.7 \% of the participants are already 
customers of at least one DTC genetic testing company and further 30.1 \% of them plan on becoming one in the future. 29.7 \% don't plan on 
becoming a DTC customer. There is no significant difference in the usage of DTC companies between chromosomal sexes (Somers-d).

@@

The openSNP project (http://opensnp.org) is an open, crowdsourced online platform for both DTC customers and researchers. 
DTC customers are encouraged to share their genotyping results and connect them with phenotypes to enable GWAS for researchers,
but the data can also be used for other research and in any other context.  
Users can create a personal profile, discuss SNPs and phenotypes on the 
platform using a simple commenting system, or send each other private messages.

\subsubsection*{Sharing genotypic information}
Currently users can upload their genotypings from the companies 23andme, deCODEme and FamilyTreeDNA via a webinterface to the openSNP 
project. The uploaded data is published under the Creative Commons Zero-license, 
which - in accordance with the Panton Principles (doi:10.1371/journal.pbio.1001195) - 
allows a complete reuse of the data without any constraints.
Between the start of openSNP on 09/27/2011 and 12/18/2012, 214 people have signed 
up with openSNP, and 79 genotyping files were made available. The openSNP 
database lists 69,486,471 SNPs, which are distributed over 1,938,603 unique SNP-IDs.
Figure n depicts the increase of users and genotyping files over time.

\subsubsection*{Crowdsourcing phenotypes}
Beside uploading genotypic data, users may create new phenotypes that are not yet 
listed by openSNP. 
The specification of these phenotypes is open and not limited 
to pre-defined categories. To reduce the amount of manual data curation, openSNP tries to harmonize 
the expression and spelling of the same phenotype or variation. We implemented an 
autocompletion-feature, which helps users to reuse already entered phenotypes.
Users are encouraged to list as many phenotypes as possible through a simple 
achievement system, rewarding users that upload their data and enter phenotypic 
information with small badges that are shown on their profile pages.

In the same timeframe that was mentioned above, all users combined have 
entered a total of 675 variations on 47 different phenotypes with those variations being 
the different values on a given trait or phenotype. See figure n for the increase of phenotypic information over time.

The mean number of users that have entered their variations for a single phenotype 
is 14.36 (SD 12.65), the median is 10. The distribution of how many users have 
entered their data per phenotype can be seen in figure n+1. The phenotype provided 
by the most users is the eye color, which has been entered by 54 users. There are 
two phenotypes which have so far only been provided by a single user: 
The SAT Writing score and triglyceride-levels.

\subsubsection*{Connection to external services}
In order to provide users with relevant information on their respective genotypes, openSNP scans databases of the scientific literature for specific SNPs. 
A total number of 15,229 documents relevant to the SNP-IDs listed in openSNP could be found in the databases of Mendeley, the Public Library of Science and SNPedia. 
Of the primary literature, 25 \% are released in open access journals and can be accessed free of charge by every user (Figure n+2). For usability reasons, 
SNPs are ranked by the amount of information gathered by the external services. The external services themselves are ranked by how easily users can access information 
from these sources. The SNPedia entries are given the highest impact, as those are already manually curated, followed by open access publications out of 
the Public Library of Science. Lowest values are given to the Mendeley results, as those aren't necessary freely available to every user. 
An entry on SNPedia is valued 2.5 times as high as a PLoS publication and 5 times as high as a Mendeley entry.

\subsubsection*{Data access}
OpenSNP offers extensive access to the data uploaded by users. Anyone can download single genotyping files for specific users, get archives of multiple genotyping files 
grouped by phenotypic variation, or access a single download that includes all genotyping files and all phenotypes in a comma-separated table. The genetic data is also 
accessible through the Distributed Annotation System, which offers all data for specific chromosomes and specific positions on single chromosomes. 
The data is additionally available over a JSON-API, which allows users to directly access data in the JSON-format.

@@

\subsection*{Survey issues}
As the survey was taken online by voluntary participants and was mainly spread in the personal genetics community, 
the results do not reflect the general population, but overrepresent those people most likely to be interested in
a project such as openSNP: customers of DTC genetic testing companies and people with a high interest in biology. 
As we wanted to find out about interest in crowdsourcing GWAS and about how people who had purchased DTC genetic 
testing felt about sharing their data with third parties, this approach was warranted.

\subsection*{GWAS and open data}
Although prices of exome or even full genome sequencing are dropping rapidly, GWAS tend to stay considerably cheaper 
and offer a possibility of getting insight into genetic variations, for example SNPs which are linked to different traits 
on a population level.

Despite this benefit in terms of cost GWAS can only detect correlations of SNPs with those traits and don't allow 
inference on the cause for any correlation. Furthermore, GWAS need a large enough sample size, 
which has to be statistically analysed by using sound methods. 
Nevertheless, GWAS are still frequently used and new associations are found. 
(we should list some recent results here -> oxytocin maybe? we could look at the twitter-timeline of @opensnporg, i publish papers i find there).

@@

With openSNP, we have built a platform that can be used by customers of DTC genetic testing to easily share their genetic and phenotypic 
data with a wide audience, as well as by scientists and interested citizens who are looking for datasets to be used in GWAS. 
Customers of DTC genetic testing also benefit from an easy access to primary literature on SNPs and genetic variations they carry. 
While we don't have collected enough data to perform full scale GWAS yet, this might be possible in the future, as user numbers are rising.

\subsection*{Privacy and health implications and ethical considerations}
As we have found in the survey on sharing such results, many of these customers are also willing to share their results with scientists and 
the public to help scientific progress, although people who have taken DTC genetic testing are considering the privacy implications that come 
with openly sharing those results.

@@

We openly address the problem of privacy implications that come with releasing genetic data twice, during registration for openSNP and during 
the upload of the DTC genetic testing results. Users have to confirm that they have read and understood the disclaimer about possible side-effects 
that can arise by publishing their data. To further improve this process we are working on implementing informed consent-processes,
like the one by \textit{Consent for Research} (http://www.weconsent.us).

@@

The survey included demographics such as age, chromosomal sex and ethnicity of the participants. Furthermore, it included questions on their 
(planned) customership of a DTC company. If the participants already were customers, they were also asked if they were already sharing their genetic and phenotypic data. 
All participants were asked if they would share their genetical or phenotypic information with their DTC company, possible answers were "Yes", "Yes, 
but only if they did not share my medical information with anybody else" and No".
